\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R:}\\ Beyond repeated measures ANOVA and Linear Mixed Models\\ \strut \\ SUPPLEMENTARY MATERIALS: APPENDIX

### 2021-11-03 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/02-Challenges.Rmd
-Working on reviewers' comments -Reorganized folders

**In a longitudinal study, it is frequently the case that the number of observations is different across all subjects. In biomedical research, this situation can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury in animals in preclinical research) leading to what is known as "missing", "incomplete" or (more properly speaking) unbalanced data [@molenberghs2004]. The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis. This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].**

**On the other hand, LMEMs and GAMs can work with missing observations, and inferences from the model are valid when the imbalance in the observations are _missing at random_ (MAR) or _completely missing at random_ (MCAR) [@west2014; @weiss2005]. In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest [@scheffer2002]. If the data are MCAR, this means that the missingness is completely unrelated to the collected information [@potthoff2006], and in the case of MNAR the missing values are dependent on their value. For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because  the missigness is unrelated to other variables of interest. **

**However, it is worth reminding that essentially "all models are wrong" [@box1976] and that the ability of LMEMs and GAMs to work with unbalanced data does not make them immune to problems that can arise due to high rates of missing data, such as sampling bias or a drastic reduction in statistical power. Researchers must ensure that the study design is statistically sound and that measures exist to minimize missing observation rates.**

>>>>

In a longitudinal study, it is frequently the case that the number of observations is different across subjects. In biomedical research, this imbalance in sample size can be caused by reasons beyond the control of the investigator (such as dropout from patients in clinical studies and attrition or injury of animals in preclinical research) leading to what is known as "missing", "incomplete" or (more generally speaking) unbalanced data [@molenberghs2004]. The rm-ANOVA model is very restrictive in these situations as it assumes that observations exist for all subjects at every time point; if that is not the case subjects with one or more missing observations are excluded from the analysis. This is inconvenient because the remaining subjects might not accurately represent the population, and statistical power is affected by this reduction in sample size [@ma2012].

On the other hand, LMEMs and GAMs can work with missing observations, and inferences from the model are valid when the imbalance in the observations are _missing at random_ (MAR) or _completely missing at random_ (MCAR) [@west2014; @weiss2005]. In a MAR scenario, the pattern of the missing information is related to some variable in the data, but it is not related to the variable of interest [@scheffer2002]. If the data are MCAR, this means that the missingness is completely unrelated to the collected information [@potthoff2006]. Missing observations can also be _missing not at random_ (MNAR) and in the case the missing observations are dependent on their value. For example, if attrition occurs in all mice that had lower weights at the beginning of a chemotherapy response study, the missing data can be considered MAR because the missigness is unrelated to other variables of interest.

However, it is worth reminding that "all models are wrong" [@box1976] and that the ability of LMEMs and GAMs to work with unbalanced data does not make them immune to problems that can arise due to high rates of missing data, such as sampling bias or a drastic reduction in statistical power. Researchers must ensure that the study design is statistically sound and that measures exist to minimize missing observation rates.

### 2021-11-03 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/03-GAM Theory.Rmd
-Working on reviewers' comments -Reorganized folders

**To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \@ref(fig:l-q-response)C. The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time. A computational requisite is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable. Because the data has 6 unique time points, we can specify a maximum of six basis functions to create the smooth (it is important to note that is not necessary to specify an equal number of basis to the number of unique values in the independent variable; less basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data).**

>>>>

To further clarify the concept of basis functions and smooth functions, consider the simulated response for Group 1 in Figure \@ref(fig:l-q-response)C. The simplest GAM model that can be used to estimate such response is that of a single smooth term for the time effect; i.e., a model that fits a smooth to the trend of the group through time. A computational requisite is that the number of basis functions to be used to create the smooth cannot be larger than the number of unique values from the independent variable. Because the data has six unique time points, we can specify a maximum of six basis functions (including the intercept) to create the smooth (it is important to note that is not necessary to specify an equal number of basis to the number of unique values in the independent variable; less basis functions can be specified to create the smooth as well, as long as they reasonably capture the trend of the data).

### 2021-11-03 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/05-Discussion.Rmd
-Working on reviewers' comments -Reorganized folders

However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented. We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature [@vishwanath2009] using $\textsf{R}$ and the package _mgcv_[@wood2017] in Section \@ref(longitudinal-GAMs), while a basic workflow for model selection is in the Appendix. One of the features of GAMs is that they go beyond a mere _p-value_ to indicate differences between groups, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure  \@ref(fig:plot-pairwise-comp) indicate. The model is therefore able to provide an estimate of significant change between the groups at time points were data was not directly measured even with missing data exists ( $\approx$ day 3 in Figure \@ref(fig:plot-pairwise-comp) A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.

>>>>

However, from a practical standpoint is equally important to demonstrate how GAMs are computationally implemented. We have provided an example on how GAMs can be fitted using simulated data that follows trends reported in biomedical literature [@vishwanath2009] using $\textsf{R}$ and the package _mgcv_[@wood2017] in Section \@ref(longitudinal-GAMs), while a basic workflow for model selection is in the [Appendix](#workflow). One of the features of GAMs is that their Bayesian interpretation allows to indicate differences between groups without the need of a _p-value_, and in turn provide a time-based estimate of shifts in the response that can be directly tied to biological values as the pairwise smooth comparisons in Figure  \@ref(fig:plot-pairwise-comp) indicate. The model is therefore able to provide an estimate of significant change between the groups at time points were data was not directly measured even with missing data exists ( $\approx$ day 3 in Figure \@ref(fig:plot-pairwise-comp) A, B ), which can be used by researchers as feedback on experiment design and to further evaluate important biological changes in future studies.
